DP Problems of Interest

### 2012-11-14 docs/roadmap/design.tex

Size analysis to know what storage size we require

@@

XXX: can we get rid of the "tabulate" by using the function name and saying that all rules are tabulated ?

@@

% ------------------------------------------------------------------------------------------------
\subsection{Memory management}
We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory. This can be the GPU or the FPGA internal memory. Usually the main memory is larger than device memory and can ultimately be extended by either disk or network storage.

@@

% ------------------------------------------------------------------------------------------------
\subsection{Memory layout}

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{LMS compiler stack}
\textbf{User-language}: define additional parameters for the recurrence\ul
\item Windowing (to convert non-serial into serial problems)
\item Input sizes, and alphabets (backtrack, input, cost)
\item Backtrack (implicitly by backtrack alphabet size) and cost matrices bit-sizes (cost maximum may be inferred using <<Yield size/grammar analysis>>)
\item Recurrence functions, devices available
\item What to keep in memory (cost, backtrack or both).
\ule
$\Downarrow$ Conversion (using an existing technique)

@@

\textbf{Binary program}

%\ol
%\item Make sure we encompass all the most common patterns of DP: check if we have higher dimensions or more complex formula.
%\item Let the user tune the window size if he wants to reduce non-serial to serial.
%\item Concerns separation: common architecture enables flexibility (exchange components) \ul
%	\item Block processor: CPU, GPU, FPGA, must allow variation of width and height
%	\item Memory stats computation (min, sum, ... column/line combinations): CPU, GPU
%	\item Scheduler (CPU): interleave block computation and memory statistics
%	\ule
%\item Common description \ul
%	\item Block kernel processor
%	\item Full block
%	\ule
%\item Discussion: wavefront, design similarities, polyhedral theory
%\ole

>>>>

{\color{red}
Size analysis to know what storage size we require

@@

XXX: can we get rid of the "tabulate" by using the function name and saying that all rules are tabulated ?
}

@@

% ------------------------------------------------------------------------------------------------
\subsection{Memory constraints}
We denote by \textit{device} the computational device on which the processing of the DP matrix (or of a computational block) is done and $M_D$ its memory. This can be the GPU or the FPGA internal memory. Usually the main memory is larger than device memory and can ultimately be extended by either disk or network storage.

@@

% ------------------------------------------------------------------------------------------------
\subsection{Memory layout}
{\color{red} XXX}

% ------------------------------------------------------------------------------------------------
\newpage
{\color{red} 
\subsection{LMS compiler stack}
\textbf{User-language}: define additional parameters for the recurrence\ul
\item Windowing (to convert non-serial into serial problems)
\item Input sizes, and alphabets (backtrack, input, cost)
\item Backtrack (implicitly by backtrack alphabet size) and cost matrices bit-sizes (cost maximum may be inferred using <<Yield size/grammar analysis>>)
\item Recurrence functions, devices available
\item What to keep in memory (cost, backtrack or both).
\ule
$\Downarrow$ Conversion (using an existing technique)

@@

\textbf{Binary program}
}
%\ol
%\item Make sure we encompass all the most common patterns of DP: check if we have higher dimensions or more complex formula.
%\item Let the user tune the window size if he wants to reduce non-serial to serial.
%\item Concerns separation: common architecture enables flexibility (exchange components) \ul
%	\item Block processor: CPU, GPU, FPGA, must allow variation of width and height
%	\item Memory stats computation (min, sum, ... column/line combinations): CPU, GPU
%	\item Scheduler (CPU): interleave block computation and memory statistics
%	\ule
%\item Common description \ul
%	\item Block kernel processor
%	\item Full block
%	\ule
%\item Discussion: wavefront, design similarities, polyhedral theory
%\ole

### 2012-11-14 docs/roadmap/draft.tex

% ------------------------------------------------------------------------------------------------
\newpage
\section{Introduction}
\subsection{Dynamic programming}
Dynamic programming consists of solving a problem by reusing subproblems solutions. The most famous example of dynamic programming is the Fibonacci series that is defined as 
\[F(n+1) = F(n)+F(n-1) \qquad \text{ with } F(0)=F(1)=1 \]

A typical characteristic of dynamic programming is that a subproblem solution is often reused multiple time, so it is often efficient to store intermediate results in a table to speed-up computation of larger problems.

There exist various categories of dynamic programming:\ul
\item Series, knapsack : that operates usually on a single dimension
\item Sequences alignment (matching two sequences at best), top-down grammar analysis, ...
\item Tree-related algorithms: phylogenetic trees raking, maximum tree independent set, ... 
\ule

Since the first category is inherently sequential and the third category is hard to parallelize efficiently, we focus on the second type of problems, which is also the most common.

{\color{red}
parallel tree-raking, does not share much with other algorithms (sparse version of computations, might not scale efficiently). Most common patterns are already enclosed by the above problems. Real input size is around 300K. We might want to also look at an $O(n^3)$-space-complex problem (like matching 3 strings $S,T,U$).

All the problems we consider use 2D storage matrix, their dependencies are an union of: \ul
\item Serial dependencies
\item Non-serial horizontal or vertical dependencies (1D non-serial)
\item Non-serial horizontal+vertical dependencies in the form $M_{(i,j)} = {\rm op}_k f ( M_{(i,k)}, M_{(k,j)} ) $
\item We have not found other type of dependencies in the literature
\ule}

\subsection{Scala and LMS}

% ------------------------------------------------------------------------------------------------
\input{problems}
\input{design}
\input{implementation}
\input{benchmarks}
\section{Future work}
\section{Conclusion}

% ------------------------------------------------------------------------------------------------
\newpage
\section{Planning/draft}
\subsubsection*{Steps}\ul
\item Re-implement vanilla combinators with lists
\item Add following problems and features:\ul
	\item Polygon triangulation (cyclic problem)
	\item SWat with arbitrary gap (multi-track grammar)
	\item Zucker (rules of different complexity $O(n^2) / O(n)$)
	\item Rules normalization
	\item Rules dependency
	\item Automatic backtracking
	\ule
\item Test / proof that parsers are correct
\item Automate test to compare against implementation
\item Build bottom-up vanilla version\ul
	\item Separate initialization (terminals) and non-terminals (speed up?)
	\item In parallel implement the code generator for CUDA, hardcode in text function bodies for CUDA
	\ule
\item LMS implementation: from normal to Rep/Exp\ul
	\item Re-implement bottom-up vanilla into LMS
	\item Re-implement CUDA version and use functions body to infer CUDA code
	\ule
\item Write report
\item Further work\ul
	\item Serial for problems larger than memory, use hybrid (Myers and Miller's algorithm with $\log_k$ with $k$ depending on available memory) approach depending available memory
	\item 
	\ule
\ule

@@

\subsubsection*{Todo @TCK}
\begin{verbatim}
Look at string templating. """xxx $var xxx"""
- Cleanup LMS CUDA generator
- CPU+GPU implementation for serial (NS?) problems larger than device memory

2. separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)

Plan:
We need to extend the language to support all 3 cases
- Parallelogram is triangular + cyclic => add a cyclic keyword that applies to the whole problem
- Square: what is the transform being done in ADP extension?

1. optimize: terminals of bounded yield + binary decomposition if possible
2. transform into: axioms (init/fixed) + rules (iterations)
3. transform to code
     axiom(i := k(i)) -> M[i,i]=k(i)
     rule(x~y := h(x,y)) -> M[i,j]=forall_i<k<j h(M[i,k],M[k,j]) for both unbounded yield
     rule(x~y_l := h(x,y)) -> M[i,j]=forall_j-l<k<j h(M[i,k],M[k,j]) for y bounded by l
     rule(x~y_l := h(x,y)) -> M[i,j]=h(M[i,j-l],M[j-l,j]) for y of exactly l

@@

\subsubsection*{Todo @Manohar}
\begin{verbatim}
- generate chain matrix recurrences to generate implementation
- more theory behind what we want to support, how we encompass all cases, ...
\end{verbatim}

\subsection*{Plan} \ol

\item \textbf{User facing language:} goals are flexibility and compactness.\ul
	\item User-facing language should be similar to related paper \cite{adp_gpu} or \href{http://hackage.haskell.org/package/ADPfusion}{\it ADP fusion}. We want to reuse the transformation mapping (problem description) $\mapsto$ (kernel implementation) for a single element.
	\item We also may want to try to make implicit transformation for code like \\
		{\tt @DP def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)}.
	\item Windowing: the user should be able to force a windowing (i.e. force a non-serial problem to be a $k$-polyadic serial problem).	
	\item 3 different cases: we care about backtrack cost or both.
	\item Backtracking: create an operator that produces the whole backtrack sequence indices.
	\ule

$\implies$ \emph{end of October}.

\item \textbf{Prototyping:} get a prototype to understand difficulties and share common base. \\ Implement a working prototype of \nameref{aswat} on CPU (for correctness), and specific platform (CUDA/FPGA). This will give us an idea of how to implement the general case. We also need to benchmark and compare both implementations to see how we compare to existing implementations and see the direction to take (which decide is faster and by how much). Here we aim to do as good an implementation for the specific platform (CPU/GPA) as possible.\\
$\implies$ \emph{end of October}.

\item \textbf{Baseline:} Also use benchmarks provided by existing implementations as baselines.\\
$\implies$ \emph{end of October}.

\item \textbf{Formalize IR:} describe the intermediate representation, formalize the framework provided to the code generators (i.e. memory management, ...).
\item \textbf{Full compiler stack:} enrich the compiler stack from both top-down (translate best user-facing language parsers) and bottom-up (parametric code generators), core of the work.
\item \textbf{Benchmark:} make sure implementations are correct, compare them other papers.
\item \textbf{Optimizations:} improve as much as possible / as long as time permits
\ole

\begin{verbatim}
ideas.txt
---------------------------------------------------
Plan:
1. Define the main problem in focus and its class
   - Dimension
   - Serial/non-serial
   - Monadic/polyadic
2. Define some other problems that belong to this class or simpler classes
3. See the feasibility for FPGA (possibly helped by CPU)
4. Implement this algorithm in both CUDA and FPGA
   -> specify all parameters so that we can compare both implementations
5. Find a common IR for the problems
6. Write a code generator for both architectures from IR
   -> make sure all the test-DP are compiled well from IR
7. Find an expressive and compact language for the user-facing representation

Translation into C++:
http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/

----------------------------------

Ideas for the language:
- we want maximum flexibility
- backtrack can be custom, we may want just to give back indices
- two problems: we care about cost and we care about whole backtrack.
	@DP def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)
  should be translated into the more complete syntax that allows more flexibility
  => get inspiration from papers to write an internal language

- we may want to keep track of indices but we do not want to give access to the backtrack matrix / cost matrix

Language side:
- define a simple way to represent some DP programs
- define a comprehensive way to define them => micro-syntax
- define the mapping into IR (actually the IR might BE the micro-syntax)
- define optimizations to IR
  - making serializable by using aggregation functions/transformations
  - avoiding the cost matrix by moving it into the wavefront
- scheduling (with memory loads for non-serializable)
- code generation for both platforms
  - possibly choice to use which platform for what part (load/compute @ cpu/gpu/fpga)

----------------------------------

Core function F:
- in: s,t strings
- in: neighbor costs: top, left, top+left
- in: neighbor stats: top, left, top+left
- out: backtrack information
- out: cost(i,j)
- out: stats

To be most efficient, we need to get exact bit-sizes of
- strings' char
- costs: min/max values
- stats: min/max values
- backtrack

=> we might want to parallelize processing within a thread that operate on 32 or 64 bit inputs (to reduce memory accesses)

@@

For stats (aggregation of the table): we need to maintain both an horizontal and vertical front
For backtracking we need to maintain the whole table
     .. | . |
	..A | B | min/max: Min(D)=Min(Min(C),Min(B))
	----+---+ sum    : Sum(D)=Sum(C)+Sum(B)-Sum(A)
	..C | D | avg    : sum both #cells and values then divide at appropriate cell
	----+---+
we many not need to store what's the previous cell, however, the previous
cell information (usually limited range) is much more compact than the score(32-64 bits)

Problem parameters:
- bit size (alphabet,stats,cost,backtrack)
- dimensionality (2D,3D for the matrix)
  - 3D problem example is Smith Waterman with arbitrary gap weight function.
- Summarizability of the set: for computing the stats of an element, we can limit ourselves to computing values up to a certain value
  -in SW with gap extending penalty, we can summarize easily.
  -in Nussinov, we _need_ to access all the previous row-column elements for the pairwise sum (without a window limit).

user-interface.txt
--------------------------------------
Reuse most of the syntactic elements and write it as uneducated programmer, apply annotations/additions on top of it or detect that it is DP and then converge to our system.

Smith-Waterman (from problems of interest)
--------------------------------------------

trait SWOps {
	type T
	type Cost = Int
	type Backtrack = (i,j) => Direction
	def rightGap(_ :_, c:Char) : [Cost, Backtrack]
}

def myRec = {
	def M[i,j] : Int = max {
		M[i,j-1] + gapRight,
		M[i-1, j] + gapLeft,
		M[i-1,j-1] + delta(S(i),T(j))
	}
}

def SWat(S0: List[Char], T0:List[Char]) : List[Pair[Char,Char]] = {
	def cost(s: Char, t: Char) = if (s==t) 2 else 1 // example cost function

	def DP(S: List[Char], T:List[Char]): [Score,Backtrack,Output[Pair[Char,Char]] = (S,T) match {
		case (_, Nil) => (0, stop)
		case (Nil, _) => (0, stop)
		case Si::Sr, Tj::Tr) =>
			val nw = SWat(Sr,Tr);
			val n  = SWat(Sr,T);
			val w  = SWat(S,Tr);
			val c = cost(Si,Tj);

			if (nw._1+c>=max(n,w)-d) then (nw._1+c,"NW",[Si,Tj])
			else if (n>=w) then (n._1-d, "N",[Si,_])
			else (w._1-d, "W",[_,Tj])
	}
	backtrack(DP(S0,T0))
}

- Char, Score, Backtrack, Output are LMS Rep objects so that we can easily detect
  when we need to use dynamic programming.
- backtrack() is a new function that take the Backtrack information and reconstruct
  the desired output backwards

We would unfold this program into two:
1. The dynamic programming that constructs all the solution and that is the core of the algorithm
2. The backtrack function that reconstruct the solution and can be run anywhere (Java, CPU, ...)

URLs.txt
------------------------------------
Some GPU algorithms:
http://hgpu.org/?cat=11

CUDPP libraries (but awfully big resulting binary):
http://code.google.com/p/cudpp/

13 dwarfs:
http://developer.amd.com/afds/assets/presentations/2155_final.pdf

http://tutorials.jenkov.com/java-reflection/fields.html
http://lampwww.epfl.ch/~michelou/scala/scala-reflection.html

>>>>

\input{introduction}
\input{problems}
\input{design}
\input{implementation}
\input{benchmarks}
\section{Future work} \ul
\item Serial for problems larger than memory, use hybrid (Myers and Miller's algorithm with $\log_k$ with $k$ depending on available memory) approach depending available memory
\item Annotation on recursive functions to use dynamic programming like \\
	{\tt @DynaProg def Fib(n:Int) = if (n<=2) return 1 else Fib(n-1)+Fib(n-2)}.
\ule

\section{Conclusion}

% ------------------------------------------------------------------------------------------------
\newpage
\section{Planning --- Work in progress}
\subsubsection*{Steps}\ol
\item Re-implement vanilla combinators with lists
\item Add following problems and features:\ul
	\item Polygon triangulation (cyclic problem) (parallelogram matrix)
	\item SWat with arbitrary gap (multi-track grammar) (rectangular matrix)
	\item Zucker (rules of different complexity $O(n^2) / O(n)$)
	\item Rules normalization
	\item Rules dependency
	\item Automatic backtracking
	\ule
\item Test/proof parsers are correct --- make sure implementation is correct
\item Automate test to compare against implementation
\item Build bottom-up vanilla version\ul
	\item Separate initialization (terminals) and non-terminals (speed up?)
	\item In parallel implement the code generator for CUDA, hardcode in text function bodies for CUDA
	\ule
\item LMS implementation: from normal to Rep/Exp\ul
	\item Re-implement bottom-up vanilla into LMS
	\item Re-implement CUDA version and use functions body to infer CUDA code
	\ule
\item Benchmarks --- compare also versus other papers.
\item Write report
\ole

@@

\subsubsection*{Todo @TCK}
\subsubsection*{Todo @Manohar}
\begin{verbatim}
Look at string templating. """xxx $var xxx"""
- Cleanup LMS CUDA generator

Separate initialization (base cases, atoms) and processing (rules, recurrences) => less cases to handle (no if)

@@

\subsection*{OldPlan} \ol
\item \textbf{User facing language:} similar to \cite{adp_gpu} or \cite{adp_fusion} \href{http://hackage.haskell.org/package/ADPfusion}{\it ADP fusion}. We want to reuse the transformation mapping (problem description) $\mapsto$ (kernel implementation) for a single element.
\item \textbf{Prototyping:} understand difficulties and share common base. Prototype of \nameref{aswat} on CUDA/FPGA. Give an idea of how to implement the general case. Benchmark and compare both implementations. Aim to do as good an implementation for the specific platform (CPU/GPA) as possible.
\item \textbf{Formalize IR:} describe the intermediate representation, formalize the framework provided to the code generators (i.e. memory management, ...).
\item \textbf{Full compiler stack:} enrich the compiler stack from both top-down (translate best user-facing language parsers) and bottom-up (parametric code generators), core of the work.
\ole

\begin{verbatim}
1. optimize: terminals of bounded yield + binary decomposition if possible
2. transform into: axioms (init/fixed) + rules (iterations)
- define optimizations to IR
  - making serializable by using aggregation functions/transformations
  - avoiding the cost matrix by moving it into the wavefront
- scheduling (with memory loads for non-serializable)
- code generation for both platforms
  - possibly choice to use which platform for what part (load/compute @ cpu/gpu/fpga)

----------------------------------
Core function F:
- in: s,t strings
- in: neighbor costs: top, left, top+left
- in: neighbor stats: top, left, top+left
- out: backtrack information
- out: cost(i,j)
- out: stats

we might want to pack into memory the matrix data and make threads operate on more than one cell.

@@

For stats (aggregation of the table): we need to maintain both an horizontal and vertical front
For backtracking we need to maintain the whole table
... | . |
..A | B | min/max: Min(D)=Min(Min(C),Min(B))
----+---+ sum    : Sum(D)=Sum(C)+Sum(B)-Sum(A)
..C | D | avg    : sum both #cells and values then divide at appropriate cell
----+---+
we many not need to store what's the previous cell, however, the previous
cell information (usually limited range) is much more compact than the score(32-64 bits)

----------------------------------------------------------------------------
Some GPU algorithms: http://hgpu.org/?cat=11
Translation into C++: http://bibiserv.cebitec.uni-bielefeld.de/macports/resources/download/
CUDPP libraries (but awfully big resulting binary): http://code.google.com/p/cudpp/
13 dwarfs: http://developer.amd.com/afds/assets/presentations/2155_final.pdf
http://tutorials.jenkov.com/java-reflection/fields.html
http://lampwww.epfl.ch/~michelou/scala/scala-reflection.html

### 2012-11-14 docs/roadmap/problems.tex

\newpage
\section{Dynamic programming problems}
% ------------------------------------------------------------------------------------------------
\subsection{Problems classification}
\subsubsection{Definitions}\ul
\item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
\item \textbf{Matrices:} we refer indifferently by the matrix or the matrices to all the intermediate cost- and backtrack-related informations that are necessary to solve the dynamic programming problem of interest. Matrices elements are usually denoted by $M_{(i,j)}$ ($i^{\rm th}$ line , $j^{\rm th}$ column).
\item \textbf{Computation block:} this is a part of the DP matrix (cost and or backtrack) that we want to compute. A block might be either a sub-matrix (rectangular) or a parallelogram, possibly cropped at its parent matrix boundaries.
\item \textbf{Wavefront:} the wavefront consists of all the data necessary to reconstruct a computation block of the DP matrix. It might include some previous lines/columns/diagonals as well as line-/column-/diagonal-wise aggregations (min, max, sum, ...).
\item \textbf{Delay:} we call delay the maximum distance between an element and its dependencies along column and lines (ex: recurrence $M_{(i,j)}=f\big(M_{(i-1,j)}, M_{(i-2,j-1)}\big)$ has delay 3).
\ule

@@

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Related problems}
The goal of this section is to demonstrate that our framework can accommodate with many {\color{red} (20-50)} problems that we have not considered at the design time.

@@

\subsubsection{Other problems}\ul
\item Dijkstra shortest path: we need a $E\times V$ matrix, along $E$ forall $V$ reduce its distance, problem is serial along $E$, non-serial along $V$ hence of complexity $O(|E|\cdot |V^2|)$ which is far worse than both $O(|V|^2)$ (min-priority queue) and $O(|E|+|V|\log |V|)$ (Fibonacci heap).
\item Fibonacci: this problem is serial 1D. Could be implemented using a placeholder element in one of the matrix dimension.
\item \href{http://archive.ite.journal.informs.org/Vol3No1/Sniedovich/\#dpmodel}{Tower of Hanoi}: 1D non-serial
\item \href{http://www.cs.ust.hk/mjg_lib/bibs/DPSu/DPSu.Files/KnPl81.PDF}{Knuth's word wrapping}: 1D non-serial
\item \href{http://en.wikipedia.org/wiki/Longest_increasing_subsequence#Efficient_algorithms}{Longest increasing subsequence}: serial but binary search algorithm more efficient: $O(n \log n)$.
\item \href{http://www.ccs.neu.edu/home/jaa/CSG713.04F/Information/Handouts/dyn_prog.pdf}{Coin Change}: 1D non-serial

{\color{red}
\item \href{http://en.wikipedia.org/wiki/Floyd-Warshall_algorithm}{Floyd-Warshall}: <it possible to move the $k$ external loop inside?> Serial with $n$ iterations
\item \href{http://en.wikipedia.org/wiki/Viterbi_algorithm}{Viterbi \footnotesize (hidden Markov models)}: $T$ non-serial iterations over a vector
\item \href{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}{Bellman-Ford} (finding the shortest distance in a graph)
\item \href{http://en.wikipedia.org/wiki/Earley_parser#Pseudocode}{Earley parser} (a type of chart parser)
\item \href{http://en.wikipedia.org/wiki/Maximum_subarray_problem}{Kadane maximum subarray} 1D serial, look at 
\href{http://www.cosc.canterbury.ac.nz/tad.takaoka/cats02.pdf}{Takaoka} for 2D
\item Structural alignment (MAMMOTH, SSAP), \href{http://rna.tbi.univie.ac.at/cgi-bin/RNAfold.cgi}{RNA structure prediction}.
\item \href{http://en.wikipedia.org/wiki/Recursive_least_squares_filter}{Recursive least squares}
\item \href{http://www.math.utep.edu/Faculty/pmdelgado2/courses/adv_algorithms/homework-08_anser.pdf}{Bitonic tour}
}
\ule

@@

- Independent set on a tree
- 0-1 Knapsack
=====> http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
\end{verbatim}

>>>>

\newpage
\section{Dynamic programming problems} \label{problems}
% ------------------------------------------------------------------------------------------------
\subsection{Problems classification}
\subsubsection{Definitions}\ul
\item \textbf{Alphabets:} the alphabet represent an enumeration of the possible values. Their size helps determining how many bits are required in the implementation to store the values. Alphabets must be defined for input, wavefront, cost and backtrack.
\item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
\item \textbf{Matrices:} we refer indifferently by the matrix or the matrices to all the intermediate cost- and backtrack-related informations that are necessary to solve the dynamic programming problem of interest. Matrices elements are usually denoted by $M_{(i,j)}$ ($i^{\rm th}$ line , $j^{\rm th}$ column).
\item \textbf{Computation block:} this is a part of the DP matrix (cost and or backtrack) that we want to compute. A block might be either a sub-matrix (rectangular) or a parallelogram, possibly cropped at its parent matrix boundaries.
\item \textbf{Wavefront:} the wavefront consists of all the data necessary to reconstruct a computation block of the DP matrix. It might include some previous lines/columns/diagonals as well as line-/column-/diagonal-wise aggregations (min, max, sum, ...).
\item \textbf{Delay:} we call delay the maximum distance between an element and its dependencies along column and lines (ex: recurrence $M_{(i,j)}=f\big(M_{(i-1,j)}, M_{(i-2,j-1)}\big)$ has delay 3).
\ule

@@

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Related problems}
The goal of this section is to demonstrate that our framework can accommodate with many problems that we have not considered at the design time.

@@

\subsubsection{Other problems}\ul
\item Dijkstra shortest path: we need a $E\times V$ matrix, along $E$ forall $V$ reduce its distance, problem is serial along $E$, non-serial along $V$ hence of complexity $O(|E|\cdot |V^2|)$ which is far worse than both $O(|V|^2)$ (min-priority queue) and $O(|E|+|V|\log |V|)$ (Fibonacci heap).
\item Fibonacci: this problem is serial 1D. Could be implemented using a placeholder element in one of the matrix dimension.
\item \href{http://archive.ite.journal.informs.org/Vol3No1/Sniedovich/\#dpmodel}{Tower of Hanoi}: 1D non-serial
\item \href{http://www.cs.ust.hk/mjg_lib/bibs/DPSu/DPSu.Files/KnPl81.PDF}{Knuth's word wrapping}: 1D non-serial
\item \href{http://en.wikipedia.org/wiki/Longest_increasing_subsequence#Efficient_algorithms}{Longest increasing subsequence}: serial but binary search algorithm more efficient: $O(n \log n)$.
\item \href{http://www.ccs.neu.edu/home/jaa/CSG713.04F/Information/Handouts/dyn_prog.pdf}{Coin Change}: 1D non-serial
\ule

 {\color{red}
XXX: Fill up to 20-50 related problems to see how general we are
\ul
\item \href{http://en.wikipedia.org/wiki/Floyd-Warshall_algorithm}{Floyd-Warshall}: <it possible to move the $k$ external loop inside?> Serial with $n$ iterations
\item \href{http://en.wikipedia.org/wiki/Viterbi_algorithm}{Viterbi \footnotesize (hidden Markov models)}: $T$ non-serial iterations over a vector
\item \href{http://en.wikipedia.org/wiki/Bellman-Ford_algorithm}{Bellman-Ford} (finding the shortest distance in a graph)
\item \href{http://en.wikipedia.org/wiki/Earley_parser#Pseudocode}{Earley parser} (a type of chart parser)
\item \href{http://en.wikipedia.org/wiki/Maximum_subarray_problem}{Kadane maximum subarray} 1D serial, look at 
\href{http://www.cosc.canterbury.ac.nz/tad.takaoka/cats02.pdf}{Takaoka} for 2D
\item Structural alignment (MAMMOTH, SSAP), \href{http://rna.tbi.univie.ac.at/cgi-bin/RNAfold.cgi}{RNA structure prediction}.
\item \href{http://en.wikipedia.org/wiki/Recursive_least_squares_filter}{Recursive least squares}
\item \href{http://www.math.utep.edu/Faculty/pmdelgado2/courses/adv_algorithms/homework-08_anser.pdf}{Bitonic tour}
\ule

@@

- Independent set on a tree
- 0-1 Knapsack
=====> http://www.cs.ucsb.edu/~suri/cs130b/NewDynProg.pdf
\end{verbatim}
}
